---
title: "plot_figures"
author: "Zsolt Balázs"
date: "24 4 2023"
output: html_document
---
# Scripts used to generate the figures of the manuscript
```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(reshape2)
library(ComplexHeatmap)
library(ggpubr)
library(ggrepel)
library(ggbiplot)
library(plotly)
library(ggdendro)
library(tidyr)
library(grid)
library(rstatix)
```
## Length distribution
```{r}
count <- read_tsv("length_uniqcounts.tsv")

count[is.na(count)] <- 0
ratio <- count %>% mutate_at(vars(-length), funs(./sum(.))) 
samplemap <- read_tsv("metadata.txt")

ratiolong <- melt(ratio, 
                  id.vars=c("length"),
                  variable.name="ID",
                  value.name="ratio")

ratiolong <- merge(x = ratiolong, y = samplemap, by = "ID", all.x = TRUE)
```

### Preparing Fragment length distribution heatmaps
### lists of samples in the two datasets
```{r}
ssdna <- samplemap$ID[samplemap$library=="ssDNA"]
dsdna <- samplemap$ID[samplemap$library=="dsDNA"]
```
### splitting ratio dataframe
```{r}
ssdnarat <- ratio[, (colnames(ratio) %in% ssdna)]
dsdnarat <- ratio[, (colnames(ratio) %in% dsdna)]
```
### dividing by healthy median, merging and taking log2
```{r}
ssdnarat$median <- apply(ssdnarat, 1, mean, na.rm=T)
ssdnadiv <- sweep(subset(ssdnarat, select = -c(median)),1,ssdnarat$median,'/')

dsdnarat$median <- apply(dsdnarat, 1, mean, na.rm=T)
dsdnadiv <- sweep(subset(dsdnarat, select = -c(median)),1,dsdnarat$median,'/')

logmed <- cbind(ssdnadiv, dsdnadiv)
logmed <- base::log2(logmed)
logmed <- cbind(ratio$length, logmed)
names(logmed)[names(logmed) == "ratio$length"] <- "length"
longlog <- melt(logmed, 
                id.vars=c("length"),
                variable.name="ID",
                value.name="log2ratio")
longlog <- merge(x = longlog, y = samplemap, by = "ID", all.x = TRUE)
```

```{r}
longlog$diag <- factor(longlog$diag, levels = c("Liver ca", "Lung ca", "Panc ca",
                                                "Healthy", "SLE"))
### for some reason sort() didn't sort these alphabetically
idorder <- c("BH01_frag", "BH01_orig", "IA05_frag", "IA05_orig", "IA06_frag",
             "IA06_orig", "IA07_frag", "IA07_orig", "IA08_frag", "IA08_orig",
             "IC15_frag", "IC15_orig", "IC17_frag", "IC17_orig", "IC20_frag",
             "IC20_orig", "IC49_frag", "IC49_orig", "IC50_frag", "IC50_orig",
             "IC51_frag", "IC51_orig", "IC52_frag", "IC52_orig", "IH01_frag",
             "IH01_orig", "IH02_frag", "IH02_orig")
longlog$ID <- factor(longlog$ID, levels=idorder)
```
## Plot heatmap of Fig1B
```{r}
heatmap_plot <- ggplot(longlog[longlog$length>75,],
                       aes(x = length, y = ID, fill = log2ratio)) +
  geom_tile() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000") +
  theme_classic() +
  theme(axis.text = element_text(size = 6),
        axis.title = element_text(size = 7),
        axis.title.y = element_blank(),
        strip.text.y = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        legend.key.size = unit(0.3, 'cm'),
        panel.spacing = unit(0.03, "lines")) +
  scale_x_continuous(name="Fragment length (bp)") +
  facet_grid(diag~., scales = "free_y", space = "free_y",
             labeller = label_wrap_gen(width = 5))
heatmap_plot
```

## Plot ratio of fragments between 100 and 150bp, Supplementary Fig 2
```{r}
rat150 <- as.data.frame(colSums(ratio[ratio$length<=150 & ratio$length>=100,][,-1]))
colnames(rat150) <- "ratio150"
rat150$ID <- rownames(rat150)
rat150samp <- merge(x = rat150, y = samplemap, by = "ID", all.x = TRUE)

p <- ggplot(rat150samp, aes(x=sample, y=ratio150, fill=processing)) + 
  geom_bar(position = "dodge", stat="identity") + 
  scale_fill_manual(values = c("darkviolet", "darkgreen")) +
  theme_classic() + 
  theme(legend.position = c(0.9, 0.8),
        axis.text = element_text(size = 6),
        axis.title = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.title = element_blank(),
        strip.text.x = element_text(size = 6),
        panel.spacing = unit(0.1, "lines"),
        legend.key.size = unit(0.4, 'cm'),
        legend.spacing.x = unit(0.04, 'cm')) +
  scale_y_continuous(name="Ratio of fragments < 150bp") +
  scale_x_discrete(name="Sample") +
  scale_color_manual(values = c("red", "grey", "yellow")) +
  facet_grid(.~group, scales = "free_x", space = "free_x")
p
```

## Tumor fraction, Fig1C
```{r}
tf <- read_tsv("ichor_results.tsv")

tf <- merge(x = tf, y = samplemap, by = "ID", all.x = TRUE)

ggplot(tf, aes(x=sample, y=tf, fill=processing)) + 
  geom_bar(position = "dodge", stat="identity") + 
  scale_fill_manual(values = c("darkviolet", "darkgreen")) +
  theme_classic() + 
  theme(legend.position = c(0.9, 0.8),
        axis.text = element_text(size = 6),
        axis.title = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.title = element_blank(),
        strip.text.x = element_text(size = 6),
        panel.spacing = unit(0.1, "lines"),
        legend.key.size = unit(0.4, 'cm'),
        legend.spacing.x = unit(0.04, 'cm')) +
  scale_y_continuous(name="Tumor fraction") +
  scale_x_discrete(name="Sample") +
  facet_grid(.~group, scales = "free_x", space = "free_x")
```

## LIQUORICE results Fig1D
```{r}
celltype <- read_tsv("LIQUORICE_results.tsv")
ct_wide <- spread(celltype, key = cell_type, value = signature)
ct_wide[, c(2:6)] <- scale(ct_wide[, c(2:6)])
celltype$signature <- scale(celltype$signature)
celltype <- pivot_longer(ct_wide, colnames(ct_wide[2:6]), 
                         names_to = "cell_type",
                         values_to = "signature")
celltype_merged <- merge(x = celltype, y = samplemap, by = "ID", all.x = TRUE)
ct_wide_merged <- merge(x = ct_wide, y = samplemap, by = "ID", all.x = TRUE)
# Run clustering

celltype_merged$diag <- factor(celltype_merged$diag, levels = c("Liver ca",
                                                                "Lung ca",
                                                                "Panc ca",
                                                                "Healthy",
                                                                "SLE"))

heatmap_plot <- ggplot(celltype_merged, aes(x = cell_type, y = ID, fill = signature)) +
  geom_tile() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000") +
  theme_classic() +
  theme(axis.text = element_text(size = 6),
        axis.title = element_blank(),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text.y = element_text(size = 6),
        legend.key.size = unit(0.3, 'cm'),
        panel.spacing = unit(0.03, "lines")
        #legend.key.size = unit(0.3, 'cm'),
        #legend.spacing.x = unit(0.03, 'cm')
  ) +
  facet_grid(diag~., scales = "free_y", space = "free_y",
             labeller = label_wrap_gen(width = 5))
heatmap_plot
```


